---
title: "Modeling the Impact of Social Determinants on HPV Vaccination Rates Among U.S. Youth"
subtitle: "BMIN503/EPID600 Final Project"
author: "Johanna Momole"
format: 
  html:
    fig-width: 8
    fig-height: 6
    toc: true #make table of contents
    toc-location: left
    toc-depth: 2
editor: visual
theme: cerulean
number-sections: true
embed-resources: true
execute: 
  warning: false
  message: false
---

------------------------------------------------------------------------

## Overview {#sec-overview}

HPV vaccination is a critical tool in preventing HPV-related cancers, yet disparities in vaccination coverage persist across the U.S. This project uses the CDC's dataset, *Vaccination Coverage Among Adolescents (13-17 Years)*, to identify key social determinants of health (SDOH) influencing HPV vaccination rates. The goal is to build predictive models that uncover patterns in low coverage rates and provide actionable insights for public health interventions.

-   **Goals**: Understand factors influencing HPV vaccination coverage.

-   **Data**: [*Vaccination Coverage Among Adolescents (13–17 Years)*](https://data.cdc.gov/Teen-Vaccinations/Vaccination-Coverage-among-Adolescents-13-17-Years/ee48-w5t6/about_data) dataset.

-   **Faculty Consulted**:

    -   Dr. Fuchiang Tsui: Guided the selection of relevant SDOH factors and advised on managing data complexity.

    -   Dr. Jesse Hsu: Recommended statistical modeling approaches for better analysis of vaccination disparities.

## Introduction {#sec-introduction}

Despite the effectiveness of HPV vaccines in preventing HPV-related cancers, vaccination rates remain uneven across the U.S. Factors such as income, urbanization, and access to healthcare contribute significantly to disparities in vaccination coverage. For example, rural and uninsured populations consistently show lower vaccination rates, underscoring the need to address social determinants of health (SDOH).

This project integrates public health knowledge with data science methodologies to investigate these disparities. By leveraging CDC vaccination data and applying machine learning models, this study aims to uncover actionable insights that can guide public health interventions. The interdisciplinary approach draws on public health, epidemiology, and data science expertise, reflecting the guidance provided by faculty consultants.

## Methods {#sec-methods}

**Libraries:** In this project, several R libraries are utilized to perform data manipulation, visualization, and modeling. Below is an overview of the key libraries:

#### 1. **Tidyverse**

-   **Purpose:** The `tidyverse` is a collection of R packages designed for data science. It provides tools for data import, manipulation, visualization, and more.

-   **Key Functions Used:**

    -   `read_csv()`: Reads CSV files into R.

    -   `filter()`, `select()`, `mutate()`, `summarize()`: Used for data cleaning and transformation.

    -   `ggplot2` (part of the tidyverse): A powerful library for creating high-quality plots.

#### 2. **Arrow**

-   **Purpose:** The `arrow` package is used for handling large datasets efficiently by leveraging the Apache Arrow framework. It enables fast data serialization and processing.

-   **Key Functions Used:**

    -   `read_feather()`: Reads data stored in the feather format for quick access.

-   **When It's Used:** For larger datasets, the feather format speeds up data reading and writing compared to traditional CSV or text formats.

#### 3. **Gtsummary**

-   **Purpose:** The `gtsummary` package creates professional summary tables for data exploration and presentation.

-   **Key Functions Used:**

    -   `tbl_summary()`: Produces a summary table of variables, including counts and descriptive statistics.

    -   `add_p()`: Adds p-values to compare groups in a dataset.

-   **Why It’s Important:** These tables provide a quick overview of the data and highlight differences between groups.

#### 4. **GGplot2**

-   **Purpose:** Part of the tidyverse, `ggplot2` is used for creating sophisticated, customizable visualizations.

-   **Key Functions Used:**

    -   `geom_bar()`, `geom_boxplot()`, `geom_violin()`: Create bar plots, boxplots, and violin plots.

    -   `geom_point()`, `geom_smooth()`: Produce scatter plots and add smoothing lines.

    -   `labs()`, `theme()`: Enhance plots with labels, themes, and other customizations.

#### 5. **RandomForest**

-   **Purpose:** The `randomForest` package implements the Random Forest algorithm for classification and regression tasks.

-   **Key Functions Used:**

    -   `randomForest()`: Builds a Random Forest model.

    -   `importance()`: Calculates feature importance for interpretation.

-   **Why It’s Used:** Random Forest is a robust, non-linear machine learning algorithm that handles complex relationships and interactions in the data.

#### 6. **Caret**

-   **Purpose:** The `caret` package simplifies model training and evaluation for a wide range of machine learning algorithms.

-   **Key Functions Used:**

    -   `train()`: Trains models with different algorithms and hyperparameter tuning.

    -   `confusionMatrix()`: Evaluates model performance.

-   **Why It’s Used:** It streamlines the process of building and validating models, making it easier to compare different approaches.

#### 7. **DT**

-   **Purpose:** The `DT` package creates interactive tables in HTML documents.

-   **Key Functions Used:**

    -   `datatable()`: Generates tables that allow sorting, filtering, and searching.

-   **Why It’s Used:** Interactive tables make it easier for readers to explore data in the final report.

```{r}
#| eval: false

# Install necessary packages
install.packages(c("readr", "tidyverse", "randomForest", "gtsummary", "ggplot2", "usmap"))
```

```{r}
# Load packages
library(readr)
library(tidyverse)
library(randomForest)
library(gtsummary)
library(ggplot2)
library(usmap)
```

**Data Description:** The dataset from the CDC, Vaccination Coverage Among Adolescents (13–17 Years), includes vaccination rates and sociodemographic characteristics by state and region. Key variables include:

-   `vaccination_status`: Binary variable indicating whether the individual was vaccinated.

-   `insurance_status`: Type of health insurance (e.g., private, Medicaid, uninsured).

-   `income_level`: Household income categories.

-   `urbanization`: Geographic classification (e.g., metropolitan, rural).

The dataset comprises 27,600 rows and 10 columns.

```{r}
# Load dataset
vaccination.data <- read_csv("C:/Users/Asus/Downloads/HPV Final Projects/Vaccination_Coverage_among_Adolescents.csv")

# Preview the dataset
glimpse(vaccination.data)

# Check for missing values
summary(vaccination.data)
```

**Data Preparation and Exploration:** The data will be cleaned to remove missing values and inconsistencies, and variables will be transformed where necessary. Summary statistics and visualizations (e.g., bar plots, histograms) will be used to explore patterns in vaccination rates.

**Modeling:** A random forest algorithm will be applied to identify significant predictors of HPV vaccination coverage. This method was chosen for its robustness and ability to handle non-linear relationships and interactions between predictors. Feature importance analysis will highlight the most influential variables.

## Results {#sec-results}

**Exploratory Findings:**

**Modeling Results:**

**Visuals:**

## Discussion {#sec-discussion}

## Conclusion

This the conclusion. The @sec-results can be invoked here.

## References
