---
title: "Predicting HPV Vaccination Socioeconomic Determinants Using NHANES Data (2021–2023)"
subtitle: "BMIN503/EPID600 Final Project"
author: "Johanna Momole"
format: 
  html:
    fig-width: 8
    fig-height: 6
    toc: true #make table of contents
    toc-location: left
    toc-depth: 2
editor: visual
theme: cerulean
number-sections: true
embed-resources: true
execute: 
  warning: false
  message: false
---

------------------------------------------------------------------------

## Overview {#sec-overview}

HPV vaccination is a critical public health measure to reduce HPV-related cancer risk. Despite its efficacy, vaccination rates remain inconsistent across demographic and socioeconomic groups in the U.S. This project explores factors influencing HPV vaccination rates using the NHANES dataset (2021–2023). Specifically, this study examines the effects of insurance coverage, socioeconomic status, and education levels. Predictive modeling and feature importance analyses are conducted to identify key barriers and provide actionable recommendations for public health interventions.

-   **Goals**: Understand factors influencing HPV vaccination coverage and predict vaccination outcomes using logistic regression and XGBoost.

-   **Data**: *NHANES Dataset* (2021–2023) focusing on HPV vaccination, insurance, poverty, and education levels.

-   **Faculty Consulted**:

    -   Dr. Fuchiang Tsui: Guided the selection of relevant factors and strategies to handle data complexity effectively.

    -   Dr. Jesse Hsu: Recommended statistical modeling approaches for analyzing disparities anad predicting trends.

## Introduction {#sec-introduction}

HPV vaccination significantly reduces the incidence of HPV-related cancers, yet disparities in vaccination rates persist. These disparities are influenced by multiple factors, including socioeconomic status, education, and access to health insurance. Understanding these factors is crucial for designing targeted interventions to improve coverage and reduce disparities.

This project analyzes vaccination rates among adolescents aged 13–17 years using NHANES data (2021–2023). By combining descriptive statistics and predictive modeling, the study uncovers barriers to vaccination and highlights actionable solutions to improve public health outcomes.

## Methods {#sec-methods}

### Load the Libraries:

In this project, several R libraries are utilized to perform data manipulation, visualization, and modeling. Below is an overview of the key libraries:

-   **tidyverse**: Data wrangling and visualization.
-   **nhanesA**: Fetching NHANES datasets.
-   **caret**: Model training and evaluation.
-   **xgboost**: Gradient boosting for predictive modeling.
-   **smotefamily**: Balancing imbalanced datasets using oversampling techniques.
-   **Matrix**: Converting dataset into a sparse matrix format for XGBoost modeling.
-   **MLmetrics**: Calculating performance metrics such as precision, recall, and F1-score.

```{r}
#| eval: false

# Install necessary packages
install.packages(c("tidyverse", "nhanesA", "caret", "xgboost", "smotefamily", "Matrix", "MLmetrics"))
```

```{r}
# Load packages
library(tidyverse)
library(nhanesA)
library(caret)
library(xgboost)
library(smotefamily)
library(Matrix)
library(MLmetrics)
```

### Load the Dataset:

The NHANES dataset (2021–2023) includes HPV vaccination information alongside socioeconomic and demographic variables. Key variables are:

-   **HPV Vaccine**
    -   **Variable Name**: IMQ060
    -   **SAS Label**: Received HPV vaccine (Females)
    -   **English Text**: Human Papillomavirus (HPV) vaccine is given to prevent HPV infection, cervical cancer, and other conditions caused by HPV in girls and women. It is given in 2 or 3 separate doses. {Have you/Has SP} ever received one or more doses of the HPV vaccine? (The brand names for the HPV vaccines are Cervarix, Gardasil, or Gardasil 9.)
    -   **Target:** Females only, 9–49 years.
    -   **Value Descriptions:**
        -   1: Yes (Count: 1095)
        -   2: No (Count: 1512)
        -   7: Refused (Count: 2)
        -   9: Don't know (Count: 244)
        -   .: Missing (Count: 8712)
-   **Health Insurance**
    -   **Variable Name**: HIQ011
    -   **SAS Label**: Covered by health insurance
    -   **English Text**: {Are you/Is SP} covered by health insurance or some other kind of healthcare plan? \[Include health insurance obtained through employment or purchased directly as well as government programs like Medicare and Medicaid that provide medical care or help pay medical bills.\]
    -   **Target:** Both males and females, 0–150 years.
    -   **Value Descriptions:**
        -   1: Yes (Count: 11,007)
        -   2: No (Count: 864)
        -   7: Refused (Count: 10)
        -   9: Don't know (Count: 29)
        -   .: Missing (Count: 23)
-   **Poverty Level (Total Savings/Cash Assets for Family)**:
    -   **Variable Name**: IND310
    -   **SAS Label**: Total savings/cash assets for the family
    -   **English Text**: Total savings or cash assets at this time for {you/NAMES OF OTHER FAMILY/your family}.
    -   **Target:** Both males and females, 0–150 years.
    -   **Value Descriptions:**
        -   1: Less than \$3000 (Count: 3950)
        -   2: \$3001–\$5000 (Count: 579)
        -   3: \$5001–\$10000 (Count: 485)
        -   4: \$10001–\$15000 (Count: 338)
        -   5: \$15001–\$20000 (Count: 167)
        -   77: Refused (Count: 61)
        -   99: Don't know (Count: 70)
        -   .: Missing (Count: 6283)
-   **Education Level (DMDEDUC2)**:
    -   **Variable Name**: IND310
    -   **SAS Label**: Total savings/cash assets for the family
    -   **English Text**: Total savings or cash assets at this time for {you/NAMES OF OTHER FAMILY/your family}.
    -   **Target:** Both males and females, 0–150 years.
    -   **Value Descriptions:**
        -   1: Less than \$3000 (Count: 3950)
        -   2: \$3001–\$5000 (Count: 579)
        -   3: \$5001–\$10000 (Count: 485)
        -   4: \$10001–\$15000 (Count: 338)
        -   5: \$15001–\$20000 (Count: 167)
        -   77: Refused (Count: 61)
        -   99: Don't know (Count: 70)
        -   .: Missing (Count: 6283)

```{r}

# Load NHANES datasets
immunization_data <- nhanes("IMQ_L")
insurance_data <- nhanes("HIQ_L")
poverty_data <- nhanes("INQ_L")
education_data <- nhanes("DEMO_L")

# Preview datasets
summary(immunization_data)
summary(insurance_data)
summary(poverty_data)
summary(education_data)
```

### Data Cleaning:

Data cleaning steps included merging datasets, recoding variables, and filtering for complete cases. Specific recoding involved:

-   HPV vaccine: Recoded as `1` (Yes) and `2` (No).

-   Insurance: Recoded as `1` (Covered) and `2` (Not Covered).

-   Poverty: Recoded into `1` (Low income: \<\$3,000) and `2` (Higher income: ≥\$3,000).

-   Education: Recoded as `1` (Below college) and `2` (College or above).

```{r}

# Merge data
merged_data <- list(immunization_data, insurance_data, poverty_data, education_data) |>
  reduce(left_join, by = "SEQN") |>
  select(SEQN, IMQ060, HIQ011, IND310, DMDEDUC2) |>
  rename(
    ID = SEQN,
    HPV_vaccine = IMQ060,
    Insurance = HIQ011,
    Poverty = IND310,
    Education = DMDEDUC2
  ) |>
  mutate(
    HPV_vaccine = case_when(
      HPV_vaccine == "Yes" ~ 1,
      HPV_vaccine == "No" ~ 2,
      TRUE ~ NA_real_
    ),
    Insurance = case_when(
      Insurance == "Yes" ~ 1,
      Insurance == "No" ~ 2,
      TRUE ~ NA_real_
    ),
    Poverty = case_when(
      Poverty == "Less than $3000" ~ 1,
      Poverty == "$3001- $5000" ~ 2,
      Poverty == "$5001-$10000" ~ 2,
      Poverty == "$10001-$15000" ~ 2,
      Poverty == "$15001-$20000" ~ 2,
      TRUE ~ NA_real_
    ),
    Education = case_when(
      Education == "Less than 9th Grade" ~ 1,
      Education == "9-11th grade (Includes 12th grade with no diploma)" ~ 1,
      Education == "High school graduate/GED or equivalent" ~ 1,
      Education == "Some college or AA degree" ~ 2,
      Education == "College graduate or above" ~ 2,
      TRUE ~ NA_real_
    )
  ) |> 
  filter(!is.na(HPV_vaccine) & !is.na(Insurance) & !is.na(Poverty) & !is.na(Education))

summary(merged_data)
```

## Results {#sec-results}

### **Descriptive Analysis**

**Distribution of HPV Vaccination**

The number of individuals who have not received the HPV vaccination (orange bar) is significantly higher than those who have been vaccinated (light blue bar).

```{r}

# Distribution of HPV vaccination
table(merged_data$HPV_vaccine)

# Visualize the distribution
barplot(
  table(merged_data$HPV_vaccine),
  main = "Distribution of HPV Vaccination Status",
  xlab = "HPV Vaccination",
  ylab = "Frequency",
  col = c("lightblue", "orange"),
  names.arg = c("Yes", "No")
)
```

**Health Insurance Coverage**

The majority of individuals are covered by health insurance (green bar), while a smaller portion lacks insurance coverage (pink bar).

-   **Implication:** Despite a majority having insurance, the low vaccination rates (as seen in the first graph) suggest that insurance coverage alone may not fully address barriers to HPV vaccination. Other factors such as education might also play a role.

```{r}

# Distribution of insurance status
table(merged_data$Insurance)

# Visualize health insurance coverage
barplot(
  table(merged_data$Insurance),
  main = "Health Insurance Coverage",
  xlab = "Insurance",
  ylab = "Frequency",
  col = c("#8aebca", "#f1c2ed"),
  names.arg = c("Covered", "Not Covered")
)
```

**Poverty Levels**

A majority of individuals fall under the "Less than \$3000" family savings category (lavender bar), indicating a lower socioeconomic status for a significant portion of the population.

-   **Implication:** Low socioeconomic status is a known barrier to accessing healthcare, including vaccinations. This finding aligns with the observed low vaccination rates.

```{r}

# Distribution of poverty levels
table(merged_data$Poverty)

# Visualize poverty levels
barplot(
  table(merged_data$Poverty),
  main = "Distribution of Poverty Levels",
  xlab = "Family Savings",
  ylab = "Frequency",
  col = c("#cebef8", "#fdf079"),
  names.arg = c(
    "Less than $3000",
    "More than $3000"
  )
)

```

**Education Levels**

A higher proportion of individuals fall under the "College or Above" category (yellow bar) compared to those "Below College" (orange bar).

-   **Implication**: Education appears to have a positive association with vaccination rates based on previous studies. However, despite a higher proportion of educated individuals, there is still a vaccination gap.

```{r}

# Distribution of education levels
table(merged_data$Education)

# Visualize education levels
barplot(
  table(merged_data$Education),
  main = "Distribution of Education Levels",
  xlab = "Education Levels",
  ylab = "Frequency",
  col = c("coral", "#fdf079"),
  names.arg = c(
     "Bellow College",
     "College or Above"
  )
)
```

**Summary of Insights**

-   **Health Insurance**: While most individuals have insurance, it is insufficient as a standalone factor to increase vaccination rates.

-   **Socioeconomic Status**: The high proportion of individuals in lower-income categories highlights financial barriers as a significant deterrent.

-   **Education**: While the majority are educated, those without higher education may still lack sufficient awareness of HPV vaccination benefits.

### **Predictive Analysis**

**XGBoost and logistic regression** were implemented to predict HPV vaccination outcomes. The SMOTE technique was applied to balance class distributions.

#### **XGBoost**

-   **Purpose**: To achieve high predictive accuracy and effectively model complex relationships.

-   **Strengths**:

    -   Handles nonlinear interactions between features.

    -   Robust to missing data and works well with imbalanced datasets.

    -   Provides feature importance scores, helping identify the most influential predictors.

    -   Highly efficient, making it suitable for large-scale predictive modeling tasks.

#### **Logistic Regression**

-   **Purpose**: To establish a baseline model and interpret the relationships between socioeconomic factors and HPV vaccination.

-   **Strengths**:

    -   Simple and interpretable, showing how each factor impacts vaccination likelihood.

    -   Useful for understanding key determinants and their statistical significance.

    -   Complements XGBoost by providing clarity on predictor effects.

**Data Preparation**

```{r}

# Showing data imbalance
table(merged_data$HPV_vaccine)

merged_data$HPV_vaccine <- as.factor(merged_data$HPV_vaccine)
levels(merged_data$HPV_vaccine)  # Should show exactly two levels (e.g., "1" and "2")

# Convert categorical predictors to numeric
smote_data <- merged_data |>
  mutate(
    Insurance = as.numeric(as.factor(Insurance)),
    Poverty = as.numeric(as.factor(Poverty)),
    Education = as.numeric(as.factor(Education))
  )

smote_data <- smote_data |>
  mutate(across(where(is.factor), ~ as.numeric(as.character(.))))

str(smote_data)  # Ensure all predictors are numeric/factor

# Apply SMOTE
smote_output <- SMOTE(
  X = smote_data[, -which(names(smote_data) == "HPV_vaccine")],  # Exclude target variable
  target = smote_data$HPV_vaccine,  # Target variable
  K = 5,  # Number of nearest neighbors
  dup_size = 2  # Oversampling factor
)

any(is.na(smote_data))  # Ensure no missing values

# Separate the majority and minority classes
minority_class <- smote_data |> filter(HPV_vaccine == "1")
majority_class <- smote_data |> filter(HPV_vaccine == "2")

# Oversample the minority class
oversampled_minority <- minority_class[sample(1:nrow(minority_class), nrow(majority_class), replace = TRUE), ]

# Combine the oversampled minority class with the majority class
balanced_data <- bind_rows(majority_class, oversampled_minority)
balanced_data$HPV_vaccine <- as.factor(balanced_data$HPV_vaccine)

# Verify balance
table(balanced_data$HPV_vaccine)
```

**Data Splitting**

```{r}

# Ensure all necessary variables are factors
balanced_data <- balanced_data |>
  mutate(
    HPV_vaccine = as.factor(HPV_vaccine),
    Insurance = as.factor(Insurance),
    Poverty = as.factor(Poverty),
    Education = as.factor(Education)
  )

# Split the dataset into training and testing sets (70% train, 30% test)
set.seed(123)
train_index <- sample(1:nrow(merged_data), 0.7 * nrow(balanced_data))
train_data <- balanced_data[train_index, ]
test_data <- balanced_data[-train_index, ]
```

**XGBoost Model**

```{r}

# Prepare data for XGBoost
train_matrix <- sparse.model.matrix(HPV_vaccine ~ Insurance + Poverty + Education - 1, data = train_data)
test_matrix <- sparse.model.matrix(HPV_vaccine ~ Insurance + Poverty + Education - 1, data = test_data)

train_labels <- as.numeric(train_data$HPV_vaccine) - 1  # XGBoost labels start at 0
test_labels <- as.numeric(test_data$HPV_vaccine) - 1

# Train the XGBoost model
xgb_model <- xgboost(
  data = train_matrix,
  label = train_labels,
  max_depth = 3,
  eta = 0.1,
  nrounds = 100,
  objective = "binary:logistic",
  verbose = 0
)

# Predict probabilities on test data
xgb_probs <- predict(xgb_model, test_matrix)

# Convert probabilities to class predictions
xgb_class <- ifelse(xgb_probs > 0.5, 1, 2)

# Convert predictions and actual labels to factors
xgb_class <- factor(xgb_class, levels = c(1, 2))
test_data$HPV_vaccine <- factor(test_data$HPV_vaccine, levels = c(1, 2))

# Evaluate the model
xgb_conf_matrix <- confusionMatrix(
  data = xgb_class,
  reference = test_data$HPV_vaccine,
  positive = "1"
)

# Extract metrics
xgb_accuracy <- xgb_conf_matrix$overall["Accuracy"]
xgb_precision <- xgb_conf_matrix$byClass["Precision"]
xgb_recall <- xgb_conf_matrix$byClass["Recall"]
xgb_f1_score <- xgb_conf_matrix$byClass["F1"]

# Print results
cat("XGBoost Confusion Matrix:\n")
print(xgb_conf_matrix$table)
cat("\nAccuracy: ", round(xgb_accuracy, 4), "\n")
cat("Precision: ", round(xgb_precision, 4), "\n")
cat("Recall: ", round(xgb_recall, 4), "\n")
cat("F1 Score: ", round(xgb_f1_score, 4), "\n")
```

**Logistic Regression Model**

```{r}

# Logistic Regression Model
logistic_model <- glm(
  HPV_vaccine ~ Insurance + Poverty + Education,
  family = "binomial",
  data = train_data
)

# Predict probabilities on test data
logistic_probs <- predict(logistic_model, test_data, type = "response")

# Convert probabilities to class predictions
logistic_class <- ifelse(logistic_probs > 0.5, 1, 2)

# Convert predictions and actual labels to factors
logistic_class <- factor(logistic_class, levels = c(1, 2))
test_data$HPV_vaccine <- factor(test_data$HPV_vaccine, levels = c(1, 2))

# Evaluate the model
logistic_conf_matrix <- confusionMatrix(
  data = logistic_class,
  reference = test_data$HPV_vaccine,
  positive = "1"
)

# Extract metrics
logistic_accuracy <- logistic_conf_matrix$overall["Accuracy"]
logistic_precision <- logistic_conf_matrix$byClass["Precision"]
logistic_recall <- logistic_conf_matrix$byClass["Recall"]
logistic_f1_score <- logistic_conf_matrix$byClass["F1"]

# Print results
cat("Logistic Regression Confusion Matrix:\n")
print(logistic_conf_matrix$table)
cat("\nAccuracy: ", round(logistic_accuracy, 4), "\n")
cat("Precision: ", round(logistic_precision, 4), "\n")
cat("Recall: ", round(logistic_recall, 4), "\n")
cat("F1 Score: ", round(logistic_f1_score, 4), "\n")
```

```{r}

# Compile results into a data frame
results <- data.frame(
  Model = c("XGBoost", "Logistic Regression"),
  Accuracy = c(round(xgb_accuracy, 4), round(logistic_accuracy, 4)),
  Precision = c(round(xgb_precision, 4), round(logistic_precision, 4)),
  Recall = c(round(xgb_recall, 4), round(logistic_recall, 4)),
  F1_Score = c(round(xgb_f1_score, 4), round(logistic_f1_score, 4))
)

# Display the model comparison
cat("\nModel Comparison:\n")
results

```

## Discussion {#sec-discussion}

### **Feature Importance Analysis**

```{r}

# Extract feature importance
xgb_importance <- xgb.importance(model = xgb_model)

# Visualize feature importance
xgb.plot.importance(xgb_importance, main = "Feature Importance (XGBoost)")

```

-   The feature importance is based on the gain, which measures the contribution of each feature to the model's decision-making.

-   The most important feature is:

    -   **Education2**: Plays the largest role in predicting the outcome.

-   **Poverty2** and **Insurance1**: Both contribute, but their importance is significantly lower compared to education.

```{r}

# Extract coefficients from the logistic regression model
logistic_coefficients <- summary(logistic_model)$coefficients

# Organize into a dataframe
logistic_importance <- data.frame(
  Feature = rownames(logistic_coefficients),
  Coefficient = logistic_coefficients[, "Estimate"]
)

# Sort by the absolute value of the coefficient for importance
logistic_importance <- logistic_importance[order(abs(logistic_importance$Coefficient), decreasing = TRUE), ]

# Visualize feature importance
library(ggplot2)
ggplot(logistic_importance, aes(x = reorder(Feature, Coefficient), y = Coefficient)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Feature Importance (Logistic Regression)", x = "Feature", y = "Coefficient")
```

-   The feature importance is based on the absolute values of the model coefficients.

-   The most influential features are:

    -   **Education2**: This suggests that higher education levels have a strong association with the outcome.

    -   **Insurance2**: Indicates that insurance coverage is also significant but less influential than education.

    -   **Poverty2**: Poverty levels contribute to the model but are less impactful compared to education and insurance.

## Conclusion

Socioeconomic disparities significantly impact HPV vaccination rates. Targeted public health interventions focusing on insurance access, education, and poverty alleviation can improve outcomes.

## Recommendations

#### 1. **Improve Insurance Coverage**

-   **Finding**: Having health insurance is one of the strongest predictors of HPV vaccination.

-   **Recommendations**:

    -   Expand access to free or low-cost vaccination programs for uninsured individuals, especially in underserved areas.

    -   Collaborate with private insurers and government programs (e.g., Medicaid) to reduce out-of-pocket costs for HPV vaccines.

    -   Launch public campaigns to raise awareness about insurance plans that cover HPV vaccination.

#### 2. **Address Poverty-Related Barriers**

-   **Finding**: Individuals from higher poverty levels are less likely to receive vaccinations.

-   **Recommendations**:

    -   Provide transportation vouchers or mobile vaccination units to reduce logistical barriers for low-income families.

    -   Offer school-based vaccination programs in low-income areas to improve accessibility.

    -   Create targeted financial incentives for low-income families to encourage vaccination, such as grocery vouchers or childcare subsidies.

#### 3. **Promote Education About HPV Vaccination**

-   **Finding**: Higher education levels are associated with increased vaccination rates.

-   **Recommendations**:

    -   Educate parents and caregivers about the benefits of HPV vaccination through community outreach programs, especially in areas with lower average educational attainment.

    -   Partner with schools, community centers, and faith-based organizations to disseminate accurate vaccine information.

    -   Counter vaccine misinformation through social media campaigns and trusted local influencers.

#### 4. **Leverage Data-Driven Outreach**

-   **Finding**: Education, insurance, and poverty are the most critical factors influencing vaccination rates.

-   **Recommendations**:

    -   Use predictive models like XGBoost to identify communities with the highest risk of low vaccination rates and focus interventions there.

    -   Develop localized outreach efforts using geospatial mapping of vaccination rates, poverty levels, and insurance coverage.

#### 5. **Policy Interventions**

-   Advocate for mandatory HPV vaccinations as part of school-entry requirements (with appropriate opt-out clauses for medical/religious reasons).

-   Lobby for state and federal funding for HPV vaccination campaigns, focusing on underinsured populations.

-   Establish partnerships with pediatric healthcare providers to encourage vaccine recommendations during routine check-ups.

## References

-   Han, J. J., Tarney, C. M., & Song, J. (2017). Variation in genital human papillomavirus infection prevalence and vaccination coverage among men and women in the USA. Future Oncology, 13(13), 1129–1132. <https://doi.org/10.2217/fon-2017-0147>

-   Herb, S. (2024). Relationship Between Geographic Location, Race, and Socioeconomic Status on Adolescent Male HPV Vaccination in the United States.\[Doctoral dissertation, Walden University\]. Walden Dissertations and Doctoral Studies. <https://scholarworks.waldenu.edu/dissertations/16240/>

-   Maness, S. B., & Thompson, E. L. (2019). Social Determinants of Human Papillomavirus Vaccine Uptake: An Assessment of Publicly Available Data. Public Health Reports, 134(3), 264–273. <https://doi.org/10.1177/0033354919838219>

-   Quinlan, J. D. (2021). Human Papillomavirus: Screening, Testing, and Prevention. American Family Physician, 104(2), 152–159.

-   Spencer, J. C., Brewer, N. T., Trogdon, J. G., Weinberger, M., Coyne-Beasley, T., & Wheeler, S. B. (2020). Cost-effectiveness of Interventions to Increase HPV Vaccine Uptake. Pediatrics, 146(6). <https://doi.org/10.1542/peds.2020-0395>

-   Villarroel, M. A., Galinsky, A. M., Lu, P. J., & Pingali, C. (2024). Human Papillomavirus Vaccination Coverage in Children Ages 9-17 Years: United States, 2022. NCHS Data Brief, 495, 1–8.

-   Xiong, S., Humble, S., Barnette, A., Brandt, H., Thompson, V., Klesges, L. M., & Silver, M. I. (2024). Associations of geographic-based socioeconomic factors and HPV vaccination among male and female children in five US states. BMC Public Health, 24(1), 1–8. <https://doi.org/10.1186/s12889-024-18206-5>

-   Yang, H., Xie, Y., Guan, R., Zhao, Y., Lv, W., Liu, Y., Zhu, F., Liu, H., Guo, X., Tang, Z., Li, H., Zhong, Y., Zhang, B., & Yu, H. (2022). Factors affecting HPV infection in U.S. and Beijing females: A modeling study. Frontiers in Public Health, 10. <https://doi.org/10.3389/fpubh.2022.1052210>
